Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-08-22 paper/zhang-cook-laa-langrene-menendez.tex
changes made as per the reviewers

{\centering \includegraphics[width=0.7\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/tour-path}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-search-1}

@@

\caption[A comparison of the searches by two optimisers]{A comparison of the searches by two optimisers: CRS (left) and SA (right) on a 2D projection problem of a six-variable dataset, \code{boa6} using holes index. Both optimisers reach the final basis with a similar index value while it takes SA longer to find the final basis. In the earlier iterations, optimisers quickly find a better basis to proceed while in the later iterations, most sampled bases fail to make an improvement on the index value and boxplot is used to summarise the distribution of the index values. There is no better basis found in the last iteration, 9 (left) and 15 (right), before reaching the maximum number of try and hence it is colored grey.}\label{fig:toy-search}
\end{figure}
\end{Schunk}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-interp-1}

@@

\caption[An inspection of the index values as the optimisation progress for two optimisers]{An inspection of the index values as the optimisation progress for two optimisers: CRS (left) and SA (right). The holes index is optimised for a 2D projection problem on the six-variable dataset \code{boa6}. Lines indicate the interpolation and dots indicate new target bases generated by the optimisers. Interpolation in both optimisation is smooth while SA is observed to first pass by some bases with higher index value before reaching the target bases in time 76-130.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

Another interesting feature to examine is the changes in the index value
between interpolating bases since the projection on these bases is shown
in the tour animation. Trace plots are created by plotting the index
value against time. Figure \ref{fig:toy-interp} presents the trace plot
of the same optimisations as Figure \ref{fig:toy-search} and one can
observe that the trace is smooth in both cases. It may seem bizarre at
first sight that the interpolation sometimes passes bases with higher
index value before it decreases to a lower target. This happens because,
on the one hand, the probabilistic acceptance in SA implies that some
worse bases will be accepted by the optimiser. In addition, the guided
tour interpolates between the current and target basis to provide a
smooth transition between projections, and sometimes a higher index
value will be observed along the interpolation path. This indicates that
a non-monotonic interpolation cannot be avoided, even for CRS. Later, in
section \ref{monotonic}, there will be a discussion on improving the
non-monotonic interpolation for CRS.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca}

@@

\caption[Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \code{boa5} using holes index]{Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. All the bases in PD have been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases.}\label{fig:toy-pca}
\end{figure}
\end{Schunk}

Apart from checking the search and progression of an optimiser, looking
at where the bases are positioned in the basis space is also of
interest. Given the orthonormality constraint, the space of projection
bases \(\mathbf{A}_{p \times d}\) is a Stiefel manifold. For
one-dimensional projections, this forms a \(p\)-dimensional sphere. A
dimensionality reduction method, e.g.~principal component analysis is
applied to first project all the bases onto a 2D space. In a projection
pursuit guided tour optimisation, there are various types of bases
involved: 1) The starting basis; 2) The search bases that the optimiser
evaluated to produce the anchor bases; 3) The anchor bases that have the
highest index value in each iteration; 4) The interpolating bases on the
interpolation path; and finally, 5) the end basis. The importance of
these bases differs and the most important ones are the starting,
interpolating, and end bases. The anchor and search bases can be turned
on with argument \texttt{details\ =\ TRUE}. Sometimes, two optimisers
can start with the same basis but finish with bases of opposite sign.
This happens because the projection is invariant to the orientation of
the basis and so is the index value. However, this creates difficulties
for comparing the optimisers since the end bases will be symmetric to
the origin. A sign flipping step is conducted to flip the signs of all
the bases in one routine if different optimisations finish at opposite
places.

Several annotations have been made to help understanding this plot. As
mentioned previously, the original basis space is a high-dimensional
sphere and random bases on the sphere can be generated via the
\CRANpkg{geozoo} \citep{geozoo} package. Along with the bases recorded
during the optimisation and a zero basis, the parameters (centre and
radius) of the 2D space can be estimated by PCA. The centre of the
circle is the first two PC coordinates of the zero matrix, and the
radius is estimated as the largest distance from the centre to the
basis. The theoretical best basis is known for simulated data and can be
labelled to compare how close the end basis found by the optimisers is
to the theoretical best. Various aesthetics, i.e.~size, alpha
(transparency), and colour, are applicable to emphasize critical
elements and adjust for the presentation. For example, anchor points and
search points are less important and hence a smaller size and alpha are
used. Alpha can also be applied on the interpolation paths to show the
start to finish from transparent to opaque.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-animated-1}

@@

An animation is another type of display to show how the search
progresses from start to finish in the space. An
\texttt{animate\ =\ TRUE} argument is used to enable the animation in
the PCA plot. Figure \ref{fig:toy-pca-animated} shows six frames from
the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional
piece of information one can learn from this animation is that CRS finds
its end basis quicker than PD since CRS finishes its search in the 5th
frame while PD is still making more progress.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-tour-1}

@@

While the previous few examples have looked at the space of 1D bases in
a unit sphere, this section visualises the space of 2D bases. Recall
that the columns in a 2D basis are orthogonal to each other, so the
space of \(p \times 2\) bases is a \(p\)-D torus. For \(p = 3\) one
would see a classical 3D torus shape as shown by the grey points in
Figure \ref{fig:toy-torus}. The two circles of the torus can be observed
to be perpendicular to each other and this can be linked back to the
orthogonality condition. Two paths from CRS and PD are plotted on top of
the torus and coloured in green and brown respectively to match the
previous plots. The final basis found by PD and CRS are shown in a
larger shape and printed below respectively:

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-torus-1}

@@

An example of non-monotonic interpolation has been given in Figure
\ref{fig:toy-interp}: a path that passes bases with higher index value
than the target one. For SA, a non-monotonic interpolation is justified
since target bases do not necessarily have a higher index value than the
current one, while this is not the case for CRS. The original trace plot
for a 2D projection problem, optimised by CRS, is shown on the left
panel of Figure \ref{fig:interruption} and one can observe that the
non-monotonic interpolation has undermined the optimiser to realise its
full potential. Hence, an interruption is implemented to stop at the
best basis found in the interpolation. The right panel of Figure
\ref{fig:interruption} shows the trace plot after implementing the
interruption and while the first two interpolations are identical, the
basis at time 61 has a higher index value than the target in the third
interpolation. Rather than starting the next iteration from the target
basis on time 65, CRS starts the next iteration at time 61 on the right
panel and reaches a better final basis.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/interruption-1}

@@

\caption[Comparison of the interpolation before and after implementing the interruption for the 2D projection problem on \code{boa6} data using holes index, optimised by CRS]{Comparison of the interpolation before and after implementing the interruption for the 2D projection problem on \code{boa6} data using holes index, optimised by CRS. On the left panel, basis with higher index value is found during the interpolation but not used. On the right panel, the interruption stops the interpolation at the basis with the highest index value for each iteration and results in a final basis with higher index value, as shown on the right panel.}\label{fig:interruption}
\end{figure}
\end{Schunk}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/polish-1}

@@

The holes index function used for all the examples before this section
produces a smooth interpolation, while this is not the case for all the
indexes An example of a noisy index function for 1D projections compares
the projected data, \(\mathbf{Y}_{n \times 1}\), to a randomly generated
normal distribution, \(\mathcal{N}_{n \times 1}\), using the Kolmogorov
test. Let \(F_{.}(n)\) be the ECDF function (empirical cummulative
distribution function), with two possible subscripts \(Y\) and
\(\mathcal{N}\) representing the projected and randomly generated data,
and \(n\) denoting the number of observation, the Normal Kolmogorov
index \(I^{nk}(n)\), is defined as:

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/noisy-better-geo-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-result-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/flip-sign-1}

@@

  \begin{itemize}
  \tightlist
  \item
    \code{explore\_trace\_search()} produces summary plots in Figure
    \ref{fig:toy-search}.
  \item
    \code{explore\_trace\_interp()} produces trace plots for the
    interpolation points in Figure \ref{fig:toy-interp}.
  \item
    \code{explore\_space\_pca()} produces the PCA plot of projection
    bases on the reduced space in Figure \ref{fig:toy-pca}. The animated
    version in Figure \ref{fig:toy-pca-animated} can be turned on via
    the argument \texttt{animate\ =\ TRUE}.
  \item
    \code{explore\_space\_tour()} produces animated tour view on the
    full space of the projection bases in Figure \ref{fig:toy-tour}.
  \end{itemize}
\item
  \code{get\_*()} extracts and manipulates certain components from the
  existing data object.

@@

\address{%
H.Sherry Zhang\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
Melbourne, Australia%
%
\\\href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}
}

\address{%
Dianne Cook\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
Melbourne, Australia%
%
\\\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}
}

\address{%
Ursula Laa\\
University of Natural Resources and Life Sciences\\%
Institute of Statistics\\
%
Vienna, Austria%
%
\\\href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}
}

\address{%
Nicolas Langrené\\
CSIRO Data61\\%
34 Village Street, Docklands VIC 3008 Australia\\
%
Melbourne, Australia%
%
\\\href{mailto:nicolas.langrene@csiro.au}{\nolinkurl{nicolas.langrene@csiro.au}}
}

\address{%
Patricia Menéndez\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
Melbourne, Australia%
%
\\\href{mailto:patricia.menendez@monash.edu}{\nolinkurl{patricia.menendez@monash.edu}}
}

>>>>

{\centering \includegraphics[width=0.7\linewidth]{/Users/hzha400/Documents/PhD/research/zzz-done/paper-tour-vis/figs/tour-path}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/zzz-done/paper-tour-vis/figs/toy-search-1}

@@

\caption[A comparison of the searches by two optimisers]{A comparison of the searches by two optimisers: CRS (left) and SA (right) on a 2D projection problem of a six-variable dataset, \code{boa6} using the holes index. Both optimisers reach the final basis with a similar index value while it takes SA longer to find the final basis. In the earlier iterations, optimisers quickly find a better basis to proceed while in the later iterations, most sampled bases fail to make an improvement on the index value and a boxplot is used to summarise the distribution of the index values. There is no better basis found in the last iteration, 9 (left) and 15 (right), before reaching the maximum number of try and hence it is colored grey. The color scale is from the customised botanical palette in the ferrn package.}\label{fig:toy-search}
\end{figure}
\end{Schunk}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/zzz-done/paper-tour-vis/figs/toy-interp-1}

@@

\caption[An inspection of the index values as the optimisation progress for two optimisers]{An inspection of the index values as the optimisation progress for two optimisers: CRS (left) and SA (right). The holes index is optimised for a 2D projection problem on the six-variable dataset \code{boa6}. Lines indicate the interpolation and dots indicate new target bases generated by the optimisers. Interpolation in both optimisation is smooth while SA is observed to first pass by some bases with higher index values before reaching the target bases in time 76-130.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

Another interesting feature to examine is the changes in the index value
between interpolating bases since the projection on these bases is shown
in the tour animation. Trace plots are created by plotting the index
value against time. Figure \ref{fig:toy-interp} presents the trace plot
of the same optimisations as Figure \ref{fig:toy-search} and one can
observe that the trace is smooth in both cases. It may seem bizarre at
first sight that the interpolation sometimes passes bases with higher
index values before it decreases to a lower target. This happens
because, on the one hand, the probabilistic acceptance in SA implies
that some worse bases will be accepted by the optimiser. In addition,
the guided tour interpolates between the current and target basis to
provide a smooth transition between projections, and sometimes a higher
index value will be observed along the interpolation path. This
indicates that a non-monotonic interpolation cannot be avoided, even for
CRS. Later, in section \ref{monotonic}, there will be a discussion on
improving the non-monotonic interpolation for CRS.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/zzz-done/paper-tour-vis/figs/toy-pca}

@@

\caption[Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \code{boa5} using holes index]{Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. The black star is the theoretical best basis the optimisers aiming to find. All the bases in PD have been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases.}\label{fig:toy-pca}
\end{figure}
\end{Schunk}

Apart from checking the search and progression of an optimiser, looking
at where the bases are positioned in the basis space is also of
interest. Given the orthonormality constraint, the space of projection
bases \(\mathbf{A}_{p \times d}\) is a Stiefel manifold. For
one-dimensional projections, this forms a \(p\)-dimensional sphere. A
dimensionality reduction method, e.g.~principal component analysis is
applied to first project all the bases onto a 2D space. In a projection
pursuit guided tour optimisation, there are various types of bases
involved: 1) The starting basis; 2) The search bases that the optimiser
evaluated to produce the anchor bases; 3) The anchor bases that have the
highest index value in each iteration; 4) The interpolating bases on the
interpolation path; and finally, 5) the end basis. The importance of
these bases differs and the most important ones are the starting,
interpolating, and end bases. Sometimes, two optimisers can start with
the same basis but finish with bases of opposite sign. This happens
because the projection is invariant to the orientation of the basis and
so is the index value. However, this creates difficulties for comparing
the optimisers since the end bases will be symmetric to the origin. A
sign flipping step is conducted to flip the signs of all the bases in
one routine if different optimisations finish at opposite places.

Several annotations have been made to help understanding this plot. As
mentioned previously, the original basis space is a high-dimensional
sphere and random bases on the sphere can be generated via the
\CRANpkg{geozoo} \citep{geozoo} package. We use PCA to estimate the 2D
space that contains all the bases recorded and the zero basis. The
centre of the 2D view is the first two PC coordinates of the zero
matrix, and the radius is estimated as the largest distance from the
centre to the basis. The theoretical best basis is known for simulated
data and can be labelled to compare how close the end basis found by the
optimisers is to the theoretical best. Various aesthetics, i.e.~size,
alpha (transparency), and colour, are applicable to emphasize critical
elements and adjust for the presentation. For example, anchor points and
search points are less important and hence a smaller size and alpha are
used. Alpha can also be applied on the interpolation paths to show the
start to finish from transparent to opaque.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/zzz-done/paper-tour-vis/figs/toy-pca-animated-1}

@@

An animation is another type of display to show how the search
progresses from start to finish in the space. Figure
\ref{fig:toy-pca-animated} shows six frames from the animation of the
PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information
one can learn from this animation is that CRS finds its end basis
quicker than PD since CRS finishes its search in the 5th frame while PD
is still making more progress.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/zzz-done/paper-tour-vis/figs/toy-tour-1}

@@

While the previous few examples have looked at the space of 1-frames in
a unit sphere, this section visualises the space of 2-frames. Recall
that the columns in a 2D basis are orthogonal to each other, so the
space of \(p \times 2\) bases is a \(p\)-D torus. For \(p = 3\) one
would see a classical 3D torus shape as shown by the grey points in
Figure \ref{fig:toy-torus}. The two circles of the torus can be observed
to be perpendicular to each other and this can be linked back to the
orthogonality condition. Two paths from CRS and PD are plotted on top of
the torus and coloured in green and brown respectively to match the
previous plots. The final basis found by PD and CRS are shown in a
larger shape and printed below respectively:

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/zzz-done/paper-tour-vis/figs/toy-torus-1}

@@

An example of non-monotonic interpolation has been given in Figure
\ref{fig:toy-interp}: a path that passes bases with a higher index value
than the target one. For SA, a non-monotonic interpolation is justified
since target bases do not necessarily have a higher index value than the
current one, while this is not the case for CRS. The original trace plot
for a 2D projection problem, optimised by CRS, is shown on the left
panel of Figure \ref{fig:interruption} and one can observe that the
non-monotonic interpolation has undermined the optimiser to realise its
full potential. Hence, an interruption is implemented to stop at the
best basis found in the interpolation. The right panel of Figure
\ref{fig:interruption} shows the trace plot after implementing the
interruption and while the first two interpolations are identical, the
basis at time 61 has a higher index value than the target in the third
interpolation. Rather than starting the next iteration from the target
basis on time 65, CRS starts the next iteration at time 61 on the right
panel and reaches a better final basis.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/zzz-done/paper-tour-vis/figs/interruption-1}

@@

\caption[Comparison of the interpolation before and after implementing the interruption for the 2D projection problem on \code{boa6} data using holes index, optimised by CRS]{Comparison of the interpolation before and after implementing the interruption for the 2D projection problem on \code{boa6} data using holes index, optimised by CRS. On the left panel, basis with a higher index value is found during the interpolation but not used. On the right panel, the interruption stops the interpolation at the basis with the highest index value for each iteration and results in a final basis with a higher index value, as shown on the right panel.}\label{fig:interruption}
\end{figure}
\end{Schunk}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/zzz-done/paper-tour-vis/figs/polish-1}

@@

The holes index function used for all the examples before this section
produces a smooth interpolation, while this is not the case for all the
indexes. An example of a noisy index function for 1D projections
compares the projected data, \(\mathbf{Y}_{n \times 1}\), to a randomly
generated normal distribution, \(\mathcal{N}_{n \times 1}\), using the
Kolmogorov test. Let \(F_{.}(n)\) be the ECDF function (empirical
cummulative distribution function), with two possible subscripts \(Y\)
and \(\mathcal{N}\) representing the projected and randomly generated
data, and \(n\) denoting the number of observation, the Normal
Kolmogorov index \(I^{nk}(n)\), is defined as:

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/zzz-done/paper-tour-vis/figs/noisy-better-geo-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/zzz-done/paper-tour-vis/figs/kol-result-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/zzz-done/paper-tour-vis/figs/flip-sign-1}

@@

  \begin{itemize}
  \tightlist
  \item
    \code{explore\_trace\_search()} produces summary plots in Figure
    \ref{fig:toy-search}.
  \item
    \code{explore\_trace\_interp()} produces trace plots for the
    interpolation points in Figure \ref{fig:toy-interp}.
  \item
    \code{explore\_space\_pca()} produces the PCA plot of projection
    bases on the reduced space in Figure \ref{fig:toy-pca}. The animated
    version in Figure \ref{fig:toy-pca-animated} can be turned on via
    the argument \texttt{animate\ =\ TRUE}.
  \item
    \code{explore\_space\_pca()} produces the PCA plot of projection
    bases on the reduced space. Figure \ref{fig:toy-pca} includes the
    additional details of anchor and search bases, which can be turned
    on by the argument \texttt{details\ =\ TRUE}. The animated version
    in Figure \ref{fig:toy-pca-animated} is produced with argument
    \texttt{animate\ =\ TRUE}.
  \item
    \code{explore\_space\_tour()} produces animated tour view on the
    full space of the projection bases in Figure \ref{fig:toy-tour}.
  \end{itemize}
\item
  \code{get\_*()} extracts and manipulates certain components from the
  existing data object.

@@

\address{%
H.Sherry Zhang\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
Melbourne, Australia\\%
%
\href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}%
}

\address{%
Dianne Cook\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
Melbourne, Australia\\%
%
\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}%
}

\address{%
Ursula Laa\\
University of Natural Resources and Life Sciences\\%
Institute of Statistics\\
%
Vienna, Austria\\%
%
\href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}%
}

\address{%
Nicolas Langrené\\
CSIRO Data61\\%
34 Village Street, Docklands VIC 3008 Australia\\
%
Melbourne, Australia\\%
%
\href{mailto:nicolas.langrene@csiro.au}{\nolinkurl{nicolas.langrene@csiro.au}}%
}

\address{%
Patricia Menéndez\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
Melbourne, Australia\\%
%
\href{mailto:patricia.menendez@monash.edu}{\nolinkurl{patricia.menendez@monash.edu}}%
}
